---
title: "PCA Analysis on. Gene Expression Data"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document presents a Principal Component Analysis (PCA) on gene expression data to assess variation across different centers.

# Setup

## Install and Load Libraries

```{r libraries}
# Install packages if not already installed
#new_packages <- c("FactoMineR", "BiocManager", "ggplot2", "dplyr", "circlize", "RColorBrewer", "readr", "psych")
#new_packages <- new_packages[!(new_packages %in% installed.packages()[,"Package"])]
#if(length(new_packages)) install.packages(new_packages)

# Load libraries
library(FactoMineR)
library(ComplexHeatmap)
library(ggplot2)
library(dplyr)
library(circlize)
library(RColorBrewer)
library(readr)
library(psych)

# BiocManager for Bioconductor packages
#if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
#BiocManager::install("ComplexHeatmap")
```

# Data Loading and Preparation

## Load Gene Expression Data

```{r load-expression-data}
# Load your expression data 
expr <- read_csv("~/BHK lab/Ravi/Ravi_Test/data/EXPR.csv")

# Display first few rows of the dataset
head(expr)
```

## Preprocess Expression Data

```{r preprocess-expression-data}
# Set row names and remove the first column
expr <- data.frame(expr, check.names = FALSE)
rownames(expr) <- expr[, 1]
expr <- expr[, -1]
```

# Principal Component Analysis

## Transpose Data for PCA

```{r transpose-data}
# Transpose the expr data frame so that samples are on rows and genes are on columns 
expr_m <- t(expr)
```

## Perform PCA

```{r perform-pca}
# Calculate PCA 
pc <- prcomp(expr_m, center = TRUE, scale. = FALSE)

# Calculate and print the percentage of variance explained by each principal component
var_res <- pc$sdev^2 / sum(pc$sdev^2) * 100
var_res <- round(var_res, 2)  # Round to 2 decimal places
print(var_res)
```

# Clinical Data Processing

## Load Clinical Data

```{r load-clinical-data}
# Load clinical data
CLIN <- read.csv("~/BHK lab/Ravi/Ravi_Test/data/CLIN.csv")
CLIN <- CLIN[, c("patient","Institution")]
```

## Merge PCA and Clinical Data

```{r merge-data}
# Merge PCA results with clinical data
pcx <- data.frame(pc$x)
pcx$patient <- rownames(pcx)
pcx_merge <- merge(pcx, CLIN, by="patient")
rownames(pcx_merge) <- pcx_merge[,1]
pcx_merge <- pcx_merge[,-1] # institution column is added
```

# Visualization

## Bar Plot

```{r scree-plot}
#To find the largest PC of data
# Create a bar plot
barplot(var_res, main="Bar Plot", xlab="Principal Component", ylab="Percentage of Variance Explained", col="skyblue")
```

## PCA Results Plot

```{r pca-results-plot}
# Create labels for the plot
xlab_text <- paste("PC1 (", var_res[1], "%)", sep = "")
ylab_text <- paste("PC2 (", var_res[2], "%)", sep = "")

# Plot PCA results
ggplot(pcx_merge, aes(PC1, PC2, color = Institution)) +
  theme_bw() +
  geom_point() +
  labs(x = xlab_text, y = ylab_text)
```


