---
title: "QC Analysis of ICB_Ravi mae"
author: "Nasim"
date: "2023-11-07"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1.PCA

Principal Component Analysis (PCA) on gene expression data to assess variation across different centers.

# Setup

## Install and Load Libraries

```{r libraries}
# Install packages if not already installed
#new_packages <- c("FactoMineR", "BiocManager", "ggplot2", "dplyr", "circlize", "RColorBrewer", "readr", "psych", "dplyr")
#new_packages <- new_packages[!(new_packages %in% installed.packages()[,"Package"])]
#if(length(new_packages)) install.packages(new_packages)

# Load libraries
library(FactoMineR)
library(ComplexHeatmap)
library(ggplot2)
library(dplyr)
library(circlize)
library(RColorBrewer)
library(readr)
library(psych)
library(MultiAssayExperiment) 
library(limma)
library(EnhancedVolcano)
library(DT)
library(tibble)





# BiocManager for Bioconductor packages
#if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
#BiocManager::install("ComplexHeatmap")
```

# Data Loading and Preparation

## Load Gene Expression Data

```{r load-expression-data}
# Load your multiassay result and extract expression data 

mae <- readRDS("~/BHK lab/Ravi_version2/data/ICB_Ravi.rds")

expr <- assays(mae)[["expr"]]

# Display first few rows of the dataset
head(expr)
```

# Principal Component Analysis

## Transpose Data for PCA

```{r transpose-data}
# Transpose the expr data frame so that samples are on rows and genes are on columns 
expr_m <- t(expr)
```

## Perform PCA

```{r perform-pca}
# Calculate PCA 
pc <- prcomp(expr_m, center = TRUE, scale. = FALSE)

# Calculate and print the percentage of variance explained by each principal component
var_res <- pc$sdev^2 / sum(pc$sdev^2) * 100
var_res <- round(var_res, 2)  # Round to 2 decimal places

print(var_res)

```

# Clinical Data Processing

## Load Clinical Data

```{r load-clinical-data}
# extract clincal data from mae
CLIN <- data.frame(colData(mae))
CLIN <- CLIN[, c("patientid","Institution")]
```

## Merge PCA and Clinical Data

```{r merge-data}
# Merge PCA results with clinical data
pcx <- data.frame(pc$x)
pcx$patientid <- rownames(pcx)
pcx$patientid <- gsub("\\_", "-", pcx$patientid)

pcx_merge <- merge(pcx, CLIN, by="patientid")
rownames(pcx_merge) <- pcx_merge[,1]
pcx_merge <- pcx_merge[,-1] # institution column is added

```

# Visualization

## Bar Plot

```{r scree-plot}
#To find the largest PC of data
# Create a bar plot
barplot(var_res, main="Bar Plot", xlab="Principal Component", ylab="Percentage of Variance Explained", col="skyblue")
```

## PCA Results Plot

```{r pca-results-plot}
# Create labels for the plot
xlab_text <- paste("PC1 (", var_res[1], "%)", sep = "")
ylab_text <- paste("PC2 (", var_res[2], "%)", sep = "")

# Plot PCA results
ggplot(pcx_merge, aes(PC1, PC2, color = Institution)) +
  theme_bw() +
  geom_point() +
  labs(x = xlab_text, y = ylab_text)

```

#2.Limma approach and Volcano plot 




```{r diff expression, include=TRUE, result = TRUE}

#removing NA in response and expression data

na_rows <- is.na(mae$response)
cleaned_response <- mae$response[!na_rows]
cleaned_expr_data <- assays(mae)[["expr"]][, !na_rows]


# 122 patient ate removing NA
design <- model.matrix(~ cleaned_response)
fit <- lmFit(cleaned_expr_data, design)
fit <- eBayes(fit)

#Top table only shows the top portion of the results
topTable(fit)
datatable(topTable(fit))


```

Preparing Data for volcano Plot 

Convert fit to Data frame and add column gene_name symbol

```{r prepare volcano, include=TRUE, result = TRUE}
result <- topTable(fit, number=Inf)  # Get all results
df <- as.data.frame(result)  # Convert to data frame

#extract gene_name from mae ( it has gene_id column and gene_name)

genedata<- data.frame(rowData(mae@ExperimentList$expr))

#subset gene_id and gene_name from the gene_data

subset_genedata <- genedata[, c("gene_id","gene_name" )]

#mergeing process

#add a gene_id column to result 
result$gene_id <- rownames(result)
#merge result and subset_genedata  by gene_id 

merge_result <- merge(result, subset_genedata, by= "gene_id")

datatable(merge_result)

```

Volcano Plot1. 

plot Volacnot plot based on P value 

```{r volcan plot, include=TRUE, result = TRUE}



EnhancedVolcano(merge_result,
    lab = merge_result$gene_name,
    x = 'logFC',
    y = 'P.Value')

```

Volcano Plot2. 

plot Volacnot plot based on "FDR"

```{r volcanplot2, include=TRUE, result = TRUE}



EnhancedVolcano(merge_result,
    lab = merge_result$gene_name,
    x = 'logFC',
    y = 'adj.P.Val')

```
#Figure 1 of paper
- Figure 1(a):Clinical characteristics in SU2C-MARK cohort.
- Figure 1(b):BOR by PDL1 TPS category.

First 1.(a)_RNA cohort in different institutions.

```{r piechart , include=TRUE, result = TRUE}


clin <- data.frame(colData(mae))
clin_subset <-clin[,c("Institution","rna")]


# Counting the occurrences of each institution
institution_counts <- table(clin_subset$Institution)

# Creating the pie chart
pie_chart <- ggplot(clin_subset, aes(x = "", fill = factor(Institution))) +
  geom_bar(width = 1, stat = "count") +
  coord_polar("y", start = 0) +
  theme_void() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = c("#999999", "#E69F00", "#56B4E9",
                               "#009E73", "#F0E442", "#0072B2",
                               "#D55E00", "#CC79A7", "#999999",
                               "#999999", "#999999", "#999999",
                               "#999999", "#999999", "#999999"))

# Printing the pie chart
print(pie_chart)

```


Figure 1(b):BOR/recist by PDL1 TPS category.


```{r bar plot, include=TRUE, result = TRUE}


# Convert colData from MultiAssayExperiment to a data frame

data <- data.frame(colData(mae))

# Select only the necessary columns
data <- data[, c("PDL1_TPS_Description", "recist", "PDL1_TPS")]

# Assuming 'data' is your dataframe
data$PDL1_TPS_Description <- forcats::fct_explicit_na(data$PDL1_TPS_Description, na_level = "Unknown")

data$PDL1_TPS_Description <- factor(data$PDL1_TPS_Description, levels = c("<1%", "1-49%", ">50%", "Unknown"))

data$recist <- forcats::fct_explicit_na(data$recist, na_level = "NE")


library(dplyr)
# Count the occurrences of combinations of 'PDL1_TPS_Description' and 'recist'
counted_data <- count(data, PDL1_TPS_Description, recist)

# Group by 'PDL1_TPS_Description'
grouped_data <- group_by(counted_data, PDL1_TPS_Description)

# Calculate frequencies
data_summary <- mutate(grouped_data, freq = n / sum(n))

library(ggplot2)
# Plotting
ggplot(data_summary, aes(x = PDL1_TPS_Description, y = freq, fill = recist)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PDL1 TPS", y = "BOR Proportion", fill = "RECIST Category") +
  geom_text(aes(label = scales::percent(freq)), position = position_stack(vjust = 0.5)) +
  theme_minimal()


```




